---
title: "Automatic Quality of Exercise Activity Assessment"
author: "Pablo Doval"
date: "February 28, 2016"
output: html_document
---

```{r "Setup", error = FALSE, echo = FALSE, warning = FALSE, message = FALSE}
# Random seed for all the study
set.seed(1979)

# Load used libraries
library("dplyr")
library("ggplot2")
library("caret")
```

# Summary

The marked increase in adoption of wearable technologies has led to quite a considerable ammount of applications and studies that focus on analyzing the quantitative aspects of the human activities tracked by these amazing devices; users of *Fitbit*, *Microsoft's Band*, etc.. are well used now to count their steps, analyze their perfomance on the track and other metrics associated to their physical activity. However, there is another less known field of study, focused on leveraging the sensor's data to infer the specific kind of activity and the quality of the training that the subject is undergoing. 

This study - which is part of the Practical Machine Learning course of the Johns Hopkins Data Sciences specialization in Coursera - will leverage the data from the Weight Lifting Exercises Dataset, obtained from [Wearable Computing: Accelerometers' Data Classification of Body Postures and Movements](http://groupware.les.inf.puc-rio.br/work.jsf?p1=10335#ixzz41Pllt7hE), and predcit, based on the movement values, the quality of the execution.


# Pre-Processing

## Data Loading 


```{r "DataLoad", echo = FALSE, warning = FALSE}

# Data loading
validationData <-read.csv(".\\Data\\pml-testing.csv", header = TRUE, na.strings = c("NA", "#DIV/0!", " "))
trainData <-read.csv(".\\Data\\pml-training.csv", header = TRUE, na.strings = c("NA", "#DIV/0!", " "))

# Get some basic statistics for simplifying the report content creation in R Markdown
totalValidationDataObservations <- nrow(validationData)
totalTrainDataObservations <- nrow(trainData)
totalValidationDataColumns <- ncol(validationData)
totalTrainDataColumns <- ncol(trainData)
numSubjects <- nrow(trainData %>% distinct(user_name))
```

Data captured using accelerometers on the belt, forearm, arm, and dumbell of 6 participants performing the same exercise - *Unilateral Dumbbell Biceps Curl* - with five different execution qualities is available at the article referenced above, and will be used for this study. The data available is split in two data sets:

 - *Train Data:*
 
      This is a set of experimental data containing **`r totalTrainDataObservations` observations** of **`r numSubjects` subjects**, with **`r totalTrainDataColumns` columns** for each of the observations. The field **classe** is the feature we aim to predict, being a feature with the following possible values:
      
      - Class A: Exactly according to the specification.
      - Class B: Throwing the elbows to the front.
      - Class C: Lifting the dumbbell only halfway.
      - Class D: Lowering the dumbbell only halfway.
      - Class E: Throwing the hips to the front.


 - *Validation Data:*
 
      The test data provided has a similar data structure, but does not include the field **classe**, which is the feature to predict. This data will be stored for later usage as a validation data set, and will not be part of the testing data for the different algorithms.

## Data Cleansing

A brief *exploratory data analysis* exercise on the data set was performed, which lead to a series of column removals from the original to create a more compact data frame to work with. These following list summarices these changes:

 - *Removal of Id column:* 
 
    This is an autonumeric identity value that does not bring any information to the data set, and cannot be used to infer the quality of the exercise. Hence, this value is discarded from the data set.
    
  - *Removal of all columns with NA values:*
  
    Most of the classification algorithms cannot be used when there are missing values. For that reason, this study will remove all columns that contain NA's or any other form of missing values in it. *NOTE: Another option could be to try and infer their values, but this scenario was ruled out on this example due to time constraints*
 
 - *Removal of Time Related columns:* 
 
    This is an autonumeric identity value that does not bring any information to the data set, and cannot be used to infer the quality of the exercise. Hence, this value is discarded from the data set.
 
- *Removal of Test Subject information:* 
 
    While demographics could be an indicator in siilar exercises, the fact that the test subjects have been instructed to do the exercises in all possible fashions render this unsuitable for te prediction purposes, so this column will be removed as well. 
 
```{r "DataCleansing", echo = FALSE, warning = FALSE}

# Removal of the identity column, time stamps and user related information
cleanTrainData <- trainData
cleanTrainData <- cleanTrainData[, -c(1:7)]

# Remove all columns that has any NA value
cleanTrainData <- cleanTrainData[, apply(cleanTrainData, 2, function(x) !any(is.na(x)))] 
trainCleanColumns <- ncol(cleanTrainData)
trainCleanRows <- nrow(cleanTrainData)

# Same transformations on the test data set. NOTE: here we cannot just apply the removal of any NA column,
# because the subset might not be the same than in the training data subset. Instead, we will just filter
# by the same column names than in the clean training data set.
cleanValidationData <- validationData[, names(cleanTrainData[, -53])]
validationCleanColumns <- ncol(cleanValidationData)
validationCleanRows <- nrow(cleanValidationData)
```
 
The resulting data set for training contains now **`r trainCleanColumns` columns**, while retaining **`r trainCleanRows` observations**, while the data set for testing is **`r validationCleanColumns` wide** with **`r validationCleanRows` observations**.

## Data Partitioning

```{r "DataPartitioning", echo = FALSE, warning = FALSE}

# Training data threshold
trainingThreshold <- 0.7

# Partitioning using the caret package
trainRows <- createDataPartition(cleanTrainData$classe, p = trainingThreshold, list = FALSE)
training <- cleanTrainData[trainRows,]
testing <- cleanTrainData[-trainRows,]

```

Since it has been decided to keep away the test data supplied by the data source for use as validation data, it is needed to generate an specific subset of data to use as testing data for the different algorithms that will be evaluated as part of this study. The new train data subset will be **`r paste(round(trainingThreshold*100,digits=1),"%",sep="")`** of the original training data available, while the remainder will be used as test data.

# Choosing the Model

# Results
 

